On-the-fly Algorithms for GKAT Equivalences

### 2025-01-24 main.tex
automata construction

The final piece of the puzzle is to convert any given expression into an \emph{equivalent} state in some symbolic GKAT coalgebra. 
This goal can be achieved by lifting existing constructions like derivatives and Thompson's construction~\cite{schmid_GuardedKleeneAlgebra_2021,smolka_GuardedKleeneAlgebra_2020} to the symbolic setting.
Then the correctness of the non-symbolic construction can be used to show the correctness of the symbolic construction i.e. we will prove that the lowering as shown in construction~\labelcref{cons:lowering} will yield the conventional derivative.
Other important properties, like finiteness of derivative coalgebra and the correctness of the Thompson's construction can be established using a symbolic GKAT coalgebra homomorphism from the Thompson's construction to the derivative.

We introduce some new notations for the transitions of symbolic GKAT coalgebra: for a symbolic GKAT coalgebra \(Ŝ ≜ (S, δ̂, ϵ̂)\) and state \(s ∈ S\), we will use \(s \transAcc{S}{b}\) to denote \(b ∈ ϵ̂(s)\); and use \(s \transvia{b ∣ p}_S s'\) to denote \((b, s', p) ∈ δ̂(s)\).

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the least symbolic GKAT coalgebra (ordered by point-wise subset ordering on \(ϵ̂\) and \(δ̂\)) that satisfy the rules in~\Cref{fig:derivatives-rules}.
Basically, a transition is in \(D̂\) if and only if it is derivable by the rules in~\Cref{fig:derivatives-rules}.
These rules are very close to the rules of GKAT derivative by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This is no coincidence as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis for our correctness argument.
\begin{theoremEnd}{theorem}[Correctness]\label{thm:derivative-correctness}
    The lowering of \(D̂\), denoted \(D\), is exactly the derivative defined by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics generated by the derivative coalgebra, \(⟦e⟧ = ⟦e⟧_{D} = ⟦e⟧_{D̂}\).
\end{theoremEnd}
\begin{proofEnd}
    TODO: unfold the statement.
\end{proofEnd}

@@

Another way to construct a coalgebra from a GKAT expression is via Thompson's construction. We lift the original construction to the symbolic setting.
A useful operation for Thompson's construction is the following guard operation, denoted by \(⟨B|\), where \(B\) is a set of boolean expressions: for a transition function \(δ̂\), a accepting function \(ϵ̂\), and a state \(s\),
\begin{align*}
    ⟨B|ϵ̂(s) & ≜ \{b ∧ c ∣ b ∈ B, c ∈ \hat{\epsilon}(s)\}; \\
    ⟨B|δ̂(s) & ≜ \{(b ∧ c, s', p) ∣ b ∈ B, (c, s', p) ∈ \hat{\delta}(s)\}.
\end{align*}
Besides guarding transition and acceptance with different conditions in if-statements and while-loops, the guard operator can also be used to simulate uniform continuation. Specifically, we can use \(⟨ϵ̂(s)|δ(s')\) to connect all the accepting transition of \(s\) to the dynamic \(δ(s')\).

@@

Our construction deviates from the original construction~\cite{smolka_GuardedKleeneAlgebra_2020} by using a start state \(s^* ∈ S\) instead of a start dynamics (or pseudo-state).
This choice will make the proof of~\Cref{thm:hom-thompson-derivative} slightly easier. 
However, as explained in~\Cref{sec:optimization-implementation}, our implementation uses start dynamics instead of start state to avoid unnecessary lookups.

@@

\Cref{thm:hom-thompson-derivative} has several consequences, one of the more obvious ones is that we can use the functoriality of the lowering operation to show the semantic equivalence of the start state in the Thompson's construction and the expression in derivative. 
% TODO: what is "in derivative"?

@@

Because of the surjectivity of \(h'\) in~\Cref{thm:suj-hom-thompson-derivative}, \(⟨s^*⟩_{Ŝ_{e}}\) has no fewer states than \(⟨e⟩_{D̂}\).
And, the number of states in \(Ŝ_{e}\) is greater than \(⟨s^*⟩_{Ŝ_{e}}\), which is greater than \(⟨e⟩_{D̂}\); by induction, the number of states in \(Ŝ_{e}\) is linear to the size of the input expression, therefore the number of states in \(⟨e⟩_{D̂}\) is at most linear to the size of the expression \(e\).

Although the size of the coalgebra generated by the derivative is smaller than Thompson's construction, the states in the derivative coalgebra \(D̂\) are expressions which can be more expensive to store. Computing the next transition of the derivative coalgebra can also take longer. On the other hand, the states of Thompson's construction only requires inductively going through the expression once to construct the entire coalgebra \(Ŝ_{e}\), and its states can be represented by more efficient constructs, like integers. In practice, we utilize techniques such as hashconsing and memoization to address the memory and time usage problems of derivative construction respectively.

>>>>

The final piece of the puzzle is to convert any GKAT expression into an \emph{equivalent} state in some symbolic GKAT coalgebra. 
This goal can be achieved by lifting existing constructions like derivatives and Thompson's construction~\cite{schmid_GuardedKleeneAlgebra_2021,smolka_GuardedKleeneAlgebra_2020} to the symbolic setting.
Then the correctness of the non-symbolic construction can be used to show the correctness of the symbolic construction i.e. we will prove that the lowering of symbolic derivatives will yield the conventional derivative.
Other important properties, like finiteness of derivative coalgebra and the correctness of the Thompson's construction can be established using a symbolic GKAT coalgebra homomorphism from the Thompson's construction to the derivative.

@@

The symbolic derivative coalgebra \(D̂\), with expressions as states, is the symbolic GKAT coalgebra generated by the rules in~\Cref{fig:derivatives-rules} i.e. a transition is in \(D̂\) if and only if it is derivable.
Our symbolic derivative resembles the GKAT derivative by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
This similarity is no coincidence, as our definition exactly lowers to the definition of theirs.
This fact can be proven by case analysis on the shape of the source expression, and forms a basis for our correctness argument.
\begin{theoremEnd}{theorem}[Correctness]\label{thm:derivative-correctness}
    The lowering of \(D̂\), denoted \(D\), is exactly the derivative defined by~\citeauthor{schmid_GuardedKleeneAlgebra_2021}~\cite{schmid_GuardedKleeneAlgebra_2021}.
    Therefore, the semantics of the expression is equal to the semantics generated by the derivative coalgebra, \(⟦e⟧ = ⟦e⟧_{D} = ⟦e⟧_{D̂}\).
\end{theoremEnd}
% \begin{proofEnd}
%     TODO: unfold the statement.
% \end{proofEnd}

@@

Similarly, we also lift the Thompson's construction~\cite{smolka_GuardedKleeneAlgebra_2020} to the symbolic setting.
A useful operation for Thompson's construction is the following guard operation, denoted by \(⟨B|\), where \(B\) is a set of boolean expressions: for a transition function \(δ̂\), an accepting function \(ϵ̂\), and a state \(s\),
\begin{align*}
    ⟨B|ϵ̂(s) & ≜ \{b ∧ c ∣ b ∈ B, c ∈ \hat{\epsilon}(s)\}; \\
    ⟨B|δ̂(s) & ≜ \{(b ∧ c, s', p) ∣ b ∈ B, (c, s', p) ∈ \hat{\delta}(s)\}.
\end{align*}
Besides guarding transition and acceptance with conditions for if-statements and while-loops, the guard operator can also be used to simulate uniform continuation. 
Specifically, we can use \(⟨ϵ̂(s)|δ(s')\) to connect all the accepting transition of \(s\) to the dynamic \(δ(s')\).

@@

Our construction deviates from the original construction~\cite{smolka_GuardedKleeneAlgebra_2020} by using a start state \(s^* ∈ S\) instead of a start dynamics (or pseudo-state).
This choice will make the proof of~\Cref{thm:hom-thompson-derivative} slightly easier. 
However, as explained in~\Cref{sec:optimization-implementation}, our implementation still utilizes start dynamics to avoid unnecessary lookups.

@@

\Cref{thm:hom-thompson-derivative} can be proven by unfolding the definition, and implies several properties we desire. 
One of the more obvious ones is the semantic equivalence of the start state in the Thompson's construction and the expression in the derivative coalgebra. 
% TODO: what is "in derivative"?

@@

Because of the surjectivity of \(h'\), \(⟨s^*⟩_{Ŝ_{e}}\) has no fewer states than \(⟨e⟩_{D̂}\).
Because the number of states in \(⟨s^*⟩_{Ŝ_{e}}\) is bounded by \(Ŝ_{e}\), and \(|Ŝ_{e}|\) is linear to the size of the input expression, the number of states in \(⟨e⟩_{D̂}\) is at most linear to the size of the expression \(e\).

% Although the size of the coalgebra generated by the derivative is smaller than Thompson's construction, the states in the derivative coalgebra \(D̂\) are expressions which can be more expensive to store. Computing the next transition of the derivative coalgebra can also take longer. On the other hand, the states of Thompson's construction only requires inductively going through the expression once to construct the entire coalgebra \(Ŝ_{e}\), and its states can be represented by more efficient constructs, like integers. In practice, we utilize techniques such as hashconsing and memoization to address the memory and time usage problems of derivative construction respectively.
